Stochastic Wind Power Forecasting

### 2020-01-09 LaTeX_code/paper/paper.tex
I added the latest versions of the paper, the poster and the presentation.

\maketitle
%\thispagestyle{empty}

%%%%%%%%% ABSTRACT

@@

Reliable wind power generation forecasting is crucial for applications such as the allocation of energy reserves, optimization of electricity price and Operation scheduling of conventional power plant. We propose a data driven model based on parametric Stochastic Differential Equations (SDEs) to captures real-world asymmetric dynamics of wind power forecast errors. The SDE proposed incorporates time derivative tracking of the forecast, time-dependent mean reversion parameter and an improved diffusion term. We are able to simulate future wind power production paths and confidence bands. The method is forecast technology agnostic and enables the comparison between different forecasting technologies on the basis of an information criteria. We apply the model to historical Uruguayan and French wind power production data and various forecasts on the period (2017-2018).

@@

Reliable wind power generation forecasting is crucial for the following applications [ref]:
\begin{itemize}
\item Allocation of energy reserves such as water levels in dams or oil and gas reserves.
\item Operation scheduling of non-controlable power plants.
\item Optimization of the price of electricity for different parties such as electric utilities, Transmission system operator (TSOs), Electricity Service providers (ESPs), Independent power producers (IPPs), and energy traders.
\item Maintenance planning such as that of power plants components and transmission lines.

@@

We propose to model wind power forecasts errors using parametric stochastic differential equations (SDEs) whose solution defines a stochastic process. This resultant stochastic process describes the time evolution dynamics of wind power forecast errors while capturing properties such as a correlation structure and the inherent asymmetry. Additionally, the model we propose is agnostic of the forecasting technology and serves to complement forecasting procedures by providing a data driven stochastic forecast. Hence, we are able to evaluate wind power forecasts according to their real-world performance and we are able to compare different forecasting technologies. Most notably, we are able to simulate future wind power production given a deterministic wind power forecast. Future wind power production as well as the analytical form of the proposed SDE can be used in optimal control problems involving wind power production.

Previous attempt by M\o ller et al. (M\o ller, Zugno, \& Madsen, 2016)  considered stochastic wind power forecast models based on stochastic differential equations (SDEs). Here, we propose an improved model featuring time derivative tracking of the forecast, time-dependent mean reversion, modified diffusion and non-Gaussian approximations. We apply the model to Uruguayan and french wind power forecasts together with historical wind power production data pertaining to the year 2017-2018.

@@

In this paper we present the phenomenological underlaying model in section (2) and describe the physical constraints in section (3) and how these constrains can be met. Then, in section (4), we will introduce an alternative formulation of the model in Lamperti space. In section (5), we show our parameter estimation procedure and its results in section (6). We compare alternative models in section (7) and different forecast providers in section (8).
\begin{figure}
  \includegraphics[width=40mm,scale=1]{plots/data_1516064400.pdf}
  \includegraphics[width=40mm,scale=1]{plots/data_1516323600.pdf}\\
  \caption{Two samples from the Uruguayan of 2018. Each sample compromises of two 72-hour paths. In yellow is an hourly wind power production forecast. In blue is the actual wind power production recorded in 10 minute intervals.}
  \label{fig:sample_data}
\end{figure}

@@

%-------------------------------------------------------------------------
\section{Phenomenological  Model}
We introduce the following phenomenological model. Let $X_t$ be the  wind power generation forecasts stochastic process defined by the  following parameterized stochastic differential equation (SDE),
\begin{equation}
\begin{split}
dX_t &= a(X_t; p_t, \dot{p}_t,\bm{\theta}) \ dt + b (X_t;p_t, \bm{\theta} ) \ dW_t \quad t > 0 \\
X_0 & = x_0
\end{split}
\label{main}
\end{equation}

@@

We would like to have this model closely follow wind power forecasts while ensuring that it is unbiased with respect to the forecast. It straightforward to show that $\E X_t = p_t$ where $p_t$ is a deterministic given wind power forecast in the model above.

@@

We have that the process $X_t$ mean reverts to the wind power forecast $p_t$ and tracks its the time derivative $\dot{p}_t$.  An example such mean reverting model without derivative tracking exhibits consistent lags as shown in Figure (\ref{fig:derivative_tracking}). See section (7) for comparisons.

@@

 We normalize the forecast and production data to Uruguay  installed power capacity. at the time of observation. Thus our process must be limited to the range $[0,1]$. To enforce this constraint, we our  drift and diffusion terms  must satisfy certain rules.

Let $\bm{\theta} = (\theta_0,\alpha)$. We would like that the process to follow the wind forecast, thus we choose a drift term that is mean reverting and tracks the derivative of the deterministic forecast $p_t$, which is an input to our model.
  \begin{equation}
    a(x; p_t,\bm{\theta})=  \dot{p_t} - \theta_t(x - p_t)
  \end{equation}
where $\theta_t>0$ is a time-dependent parameter that controls the speed of reversion.

@@

\section{State independent diffusion formulation:}
The model we have demonstrated previously (\ref{VtSDE}) is  a state-dependent diffusion formulation, that is the diffusion of the SDE clearly depends on the state.  We are interested in state independent diffusion as it is more tractable and numerically stable.To achieve that, we apply a Lamperti transform to obtain the following state-independent diffusion SDE,
\begin{equation}
  \begin{split}
    dZ_t&= \frac{- \theta_t (1+ \sin(Z_t) - 2p_t) + \alpha \theta_t \sin (Z_t)   }{\cos (Z_t)} \ dt + \sqrt{2 \alpha \theta_t} dW_t \\
    Z_0&=z_0
  \end{split}
\end{equation}
where the Lamperti trasform is given by $Z_t = \arcsin \left( \frac{1}{2} \left( V_t+p_t \right) - 1 \right) $. We can see in figure (\ref{fig:skew_after_lamperti} ) the effects of the Lamperti transform on the forecast error data. The Lamperti transform has greatly reduced  the forecast error skewness and ensures that the process stays in the range $[0,1]$. Thus, in this case,  we can consider using a Gaussian density as a proxy transition density.

\section{Inference and Optimization}

@@

Since $V_t$ defined by our SDE is Markovian, the likelihood function can be written as a product of transition densities.  Consider a set of M paths with N observations each, $ V^{M,N}=\{ V_{t_1^{M,N}} , V_{t_2^{M,N}} ,\ldots , V_{t_N^{M,N}} \}$ observed in intervals of $\Delta_N$.

@@

\begin{equation}
\begin{split}
\frac{ \partial f }{\partial t } & (y ,t | x , s, \theta_0 , \alpha )= - \frac{\partial}{ \partial y} ( a( y;  p_t, \dot{p}_t, \theta_0 ) f( y ,t | x , s, \theta_0  , \alpha ) ) \\
& + \frac{1}{2} \frac{\partial^2}{ \partial y^2} ( b^2(y; p_t, \theta_0, \alpha  )  f(y ,t | x , s,\theta_0 , \alpha ) ) \quad  t < s\\
\end{split}
\end{equation}
This is a parametric PDE which is computationally expensive to solve and optimize for every transition. Therefore, we consider approximate likelihood methods.

@@

We propose a proxy transition density. We match the moments of our SDE model with that of the proxy density. Using It$\hat{o}$ formula, we arrive at the following iterative ODEs for the state dependent diffusion formulation (\ref{VtSDE})
\begin{equation}
\begin{split}
\frac{d \E[ V^k_t]}{dt} = - k \theta_t \E [ V^k_t] + \frac{k(k-1)}{2} \E [ V^{k-2}_t  b(V^k_t;\theta_t, \alpha)]
\end{split}
\end{equation}
For $t\in [t_{n-1}, t]$, the first two moments are given by
\begin{equation}
\begin{split}
\frac{d m_1 (t)}{dt} &= - m_1(t)\theta_t \\
\frac{d m_2 (t)}{dt} &=  -2 m_2(t)\theta_t(1+\alpha) + 2\alpha\theta_t m_1(t)(1-2p_t) \\
&\quad + 2 \alpha\theta_t p_t (1-p_t)\\
\end{split}
\end{equation}
with initial conditions, $m_1(t_{n-1})= v_{n-1}$ and $m_2(t_{n-1})= v_{n-1}^2$.

@@

To initialize the the optimization process, we solve the following least-squares problem which gives us first estimates of the mean reversion parameter $\theta_0$.

@@

Solving for $\alpha$, we have both first estimates to kick start the inference processes given as follows:
\begin{enumerate}
\item[Step 1.]  initialize
\item[Step 2.]  optimize the log-likelihood function using  Nelder-Mead optimization algorithm on a mini-batch sampled with replacement.
\item[Step 3. ] check if  accuracy threshold is reached. Else, re-initialize the optimization with the most recent result on a larger mini-batch.
\end{enumerate}

\section{Model Comparison}
We compare the following candidate models to find the best-fit that maximizes the retained information,
\begin{itemize}
  \item Model 0: This model is the most basic model without derivative tracking.
  \begin{equation}
  \begin{split}
  dX_t &=  - \theta_0 (p_t-X_t) \  dt + \sqrt{2 \theta \alpha X_t (1-X_t)} \  dW_t  \\ %\quad t > 0
  X_0 & = x_0
\end{split}\label{M0}
  \end{equation}
  with $\theta_t= \theta_0 $.

@@

\section{Forecast Provider Comparison}
We compare forecasts from two different companies for the same period.
\begin{table}[]
\centering
\begin{tabular}{|c|l|c|c|}
\hline
Forecast Provider & parameters $(\theta_0, \alpha)$ & AIC & BIC \\ \hline
Provider A  &  $(\add,\add)$      &  \add   & \add    \\ \hline
Provider B  & $(\add,\add) $  &  \add   &   \add  \\ \hline
\end{tabular}
\caption{\add}
\label{tab:forcast_comparison}
\end{table}
In Figure (\add)-(\add), we see that forecast provider A is of better quality than provider B. This is confirmed by both the AIC and BIC information criteria.

@@

\section{Results}

@@

\section{Conclusions}

>>>>

\maketitle
%\thispagestyle{empty}
{\color{red}Keywords: Indirect inference, wind power, probabilistic forecasting, model selection, sensitivity.}
%%%%%%%%% ABSTRACT

@@

Reliable wind power generation forecasting is crucial for applications such as the allocation of energy reserves, optimization of electricity price and Operation scheduling of conventional power plant. We propose a data driven model based on parametric Stochastic Differential Equations (SDEs) to captures real-world asymmetric dynamics of wind power forecast errors. The SDE proposed incorporates time derivative tracking of the forecast, time-dependent mean reversion parameter and an improved diffusion term. We are able to simulate future wind power production paths and confidence bands. The method is forecast technology agnostic and enables the comparison between different forecasting technologies on the basis of an information criteria. We apply the model to historical French and Uruguayan wind power production data and forecasts on the period (2017-2018).

@@

Reliable wind power generation forecasting is crucial for the following applications [ref]:
\begin{itemize}
\item Allocation of energy reserves such as water levels in dams or oil, and gas reserves.
\item Operation scheduling of non-controlable power plants.
\item Optimization of the price of electricity for different parties such as electric utilities, Transmission system operator (TSOs), Electricity Service providers (ESPs), Independent power producers (IPPs), and energy traders.
\item Maintenance planning such as that of power plants components and transmission lines.

@@

We propose to model wind power forecasts errors using parametric stochastic differential equations (SDEs) whose solution defines a stochastic process. This resultant stochastic process describes the time evolution dynamics of wind power forecast errors while capturing properties such as a correlation structure and the inherent asymmetry. Additionally, the model we propose is agnostic of the forecasting technology and serves to complement forecasting procedures by providing a data driven stochastic forecast. Hence, we are able to evaluate wind power forecasts according to their real-world performance and we are able to compare different forecasting technologies. Most notably, we are able to simulate future wind power production given a deterministic wind power forecast. Future wind power production using Monte Carlo methods, as well as the analytic form of the proposed SDE, can be used in optimal control problems involving wind power production.

Previous attempt by M\o ller et al. (M\o ller, Zugno, \& Madsen, 2016)  considered stochastic wind power forecast models based on stochastic differential equations (SDEs). Here, we propose an improved model featuring time derivative tracking of the forecast, time-dependent mean reversion, modified diffusion and non-Gaussian approximations. We apply the model to French and Uruguayan wind power forecasts together with historical wind power production data pertaining to the year 2017-2018.

@@

In this paper we present the phenomenological underlying model in Section \ref{Section_2} and describe the physical constraints in Section \ref{Section_3} and how these constrains can be met. Then, in Section \ref{Section_4}, we will introduce an alternative formulation of the model in Lamperti space. In Section \ref{Section_5}, we show our parameter estimation procedure and its results in Section \ref{Section_6}. We compare alternative models in Section \ref{Section_7} and different forecast providers in Section \ref{Section_8}.
\begin{figure}
  \includegraphics[width=40mm,scale=1]{plots/data_1516064400.pdf}
  \includegraphics[width=40mm,scale=1]{plots/data_1516323600.pdf}\\
  \caption{Two samples from the Uruguayan of 2018. Each sample compromises of two 72-hour paths. In yellow is an hourly wind power production forecast. In blue is the actual wind power production recorded in 10 minute intervals.}
  \label{fig:sample_data}
\end{figure}

@@

%-------------------------------------------------------------------------
\section{Phenomenological  Model} \label{Section_2}

We introduce the following phenomenological model. Let $X_t$ be the  wind power generation forecasts stochastic process defined by the  following parameterized stochastic differential equation (SDE),
\begin{equation}
\begin{split}
dX_t &= a(X_t; p_t, \dot{p}_t,\bm{\theta}) \ dt + b (X_t;p_t, \bm{\theta} ) \ dW_t \quad t > 0 \\
X_0 & = x_0
\end{split}
\label{main}
\end{equation}

@@

We would like to have this model closely follow wind power forecasts while ensuring that it is unbiased with respect to the forecast. It straightforward to show that $\E\left[X_t\right] = p_t$, where $p_t$ is a deterministic given wind power forecast in the model above.

@@

We have that the process $X_t$ mean reverts to the wind power forecast $p_t$ and tracks its the time derivative $\dot{p}_t$.  An example such mean reverting model without derivative tracking exhibits consistent lags as shown in Figure (\ref{fig:derivative_tracking}). See Section \ref{Section_7} for comparisons.

@@

 We normalize the forecast and production data to Uruguay  installed power capacity at the time of observation. Thus our process must be limited to the range $[0,1]$. To enforce this constraint, our  drift and diffusion terms  must satisfy certain rules.

Let $\bm{\theta} = (\theta_0,\alpha)$. We want a process that follow the wind forecast, thus we choose mean reverting drift term which also tracks the derivative of $p_t$, which is an input to our model.
  \begin{equation}
    a(x; p_t,\bm{\theta})=  \dot{p_t} - \theta_t(x - p_t)
  \end{equation}
where $\theta_t>0$ is a time-dependent parameter that controls the speed of reversion.

@@

\section{State independent diffusion: Lamparti transform} \label{Section_3}

Our model in (\ref{VtSDE}) is  a state-dependent diffusion formulation, that is the diffusion of the SDE clearly depends on the state.  We are interested in state independent diffusion as it is more tractable and numerically stable. To achieve that, we apply a Lamperti transform to obtain the following state-independent diffusion SDE,
\begin{equation}
  \begin{split}
    dZ_t&= \frac{- \theta_t (1+ \sin(Z_t) - 2p_t) + \alpha \theta_t \sin (Z_t)   }{\cos (Z_t)} \ dt + \sqrt{2 \alpha \theta_t} dW_t \\
    Z_0&=z_0
  \end{split}
\end{equation}
where the Lamperti trasform is given by $Z_t = \arcsin \left( \frac{1}{2} \left( V_t+p_t \right) - 1 \right) $. We can see in figure (\ref{fig:skew_after_lamperti}) the effects of the Lamperti transform on the forecast error data. The Lamperti transform has greatly reduced  the forecast error skewness and ensures that the process stays in the range $[0,1]$. Thus, in this case,  we can consider using a Gaussian density as a proxy transition density.

\section{Likelihood in $V$ space} \label{Section_4}

@@

Since $V_t$ defined by our SDE is Markovian, the likelihood function can be written as a product of transition densities.  Consider the set of $M$ paths with $N$ observations each, $ V^{M,N}=\{ V_{t_1^{M,N}} , V_{t_2^{M,N}} ,\ldots , V_{t_N^{M,N}} \}$, observed in $\Delta_N$ intervals.

@@

\begin{equation}
\begin{split}
\frac{ \partial f }{\partial t } & (v ,t | x , s, \theta_0 , \alpha )= - \frac{\partial}{ \partial v} ( a( v;  p_t, \dot{p}_t, \theta_0 ) f( v ,t | x , s, \theta_0  , \alpha ) ) \\
& + \frac{1}{2} \frac{\partial^2}{ \partial v^2} ( b^2(v; p_t, \theta_0, \alpha  )  f(v ,t | x , s,\theta_0 , \alpha ) ) \quad  t < s\\
\end{split}
\end{equation}
This is a parametric PDE which is computationally expensive to solve and optimize for every transition. Therefore, we consider approximate likelihood methods.

@@

We propose a proxy transition density. We match the moments of our SDE model with that of the proxy density. Using It\^o formula, we arrive at the following iterative ODEs for the state dependent diffusion formulation (\ref{VtSDE})
\begin{equation}
\begin{split}
\frac{d \E[ V^k_t]}{dt} = - k \theta_t \E [ V^k_t] + \frac{k(k-1)}{2} \E [ V^{k-2}_t  b(V^k_t;\theta_t, \alpha)]
\end{split}
\end{equation}
For $t\in [t_{n-1}, t]$, the first two moments are given by
\begin{equation}
\begin{split}
\frac{d m_1 (t)}{dt} &= - m_1(t)\theta_t \\
\frac{d m_2 (t)}{dt} &=  -2 m_2(t)\theta_t(1+\alpha) + 2\alpha\theta_t m_1(t)(1-2p_t) \\
&\quad + 2 \alpha\theta_t p_t (1-p_t)\\
\end{split}
\end{equation}
with initial conditions, $m_1(t_{n-1})= v_{n-1}$ and $m_2(t_{n-1})= v_{n-1}^2$.

@@

To initialize the optimization process, we solve the following least-squares problem which gives us first estimates of the mean reversion parameter $\theta_0$.

@@

Solving for $\alpha$, we have both first estimates to quick start the inference processes given as follows:
\begin{enumerate}
\item[Step 1.]  initialize
\item[Step 2.]  optimize the log-likelihood function using  Nelder-Mead optimization algorithm on a mini-batch sampled with replacement.
\item[Step 3. ] check if  accuracy threshold is reached. Else, re-initialize the optimization with the most recent result on a larger mini-batch.
\end{enumerate}

\section{Model Comparison} \label{Section_5}
We compare the following candidate models to find the best-fit that maximizes the retained information,
\begin{itemize}
  \item Model 0: This model is the most basic model without derivative tracking.
  \begin{equation}
  \begin{split}
  dX_t &=  - \theta_0 (p_t-X_t) \  dt + \sqrt{2 \theta \alpha X_t (1-X_t)} \  dW_t  \\ %\quad t > 0
  X_0 & = x_0
\end{split}\label{M0}
  \end{equation}
  with $\theta_t= \theta_0 $.

@@

\section{Forecast Provider Comparison} \label{Section_6}
We compare forecasts from two different companies for the same period.
\begin{table}[]
\centering
\begin{tabular}{|c|l|c|c|}
\hline
Forecast Provider & parameters $(\theta_0, \alpha)$ & AIC & BIC \\ \hline
Provider A  &  $(\add,\add)$      &  \add   & \add    \\ \hline
Provider B  & $(\add,\add) $  &  \add   &   \add  \\ \hline
\end{tabular}
\caption{\add}
\label{tab:forcast_comparison}
\end{table}
In Figure (\add)-(\add), we see that forecast provider A is of better quality than provider B. This is confirmed by both the AIC and BIC information criteria.

@@

\section{Results} \label{Section_7}

@@

\section{Conclusions} \label{Section_8}

{\color{red} Application: Using this tool, it is possible to iterate over hyper parameters for the wind power forecasting, reaching an optimal forecasting.}
